Variance reduction for Markov chains via martingale representations

### 2020-05-02 aap/Revision/vr-mcmc-revision-new-new.tex
update

Now, from the H\"older inequality, we obtain (with $\|X\|_p = (\mathsf{E}X^p)^{\frac{1}{p}}$)
\begin{align*}
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=l}^{j}\alpha_{k}\left( \sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right] & \leq\sum_{j=1}^{q}\left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left\Vert \sum_{p=j+1}^{q+1} \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}^{2}\\
 & \leq\sum_{j=1}^{q} \left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left(\sum_{p=j+1}^{q+1}\left\Vert \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}\right)^{2}.
\end{align*}
Now using the fact that $\prod\limits_{k=j+1}^{p}\alpha_{k} \leq \exp\left(-\sum\limits_{k=j+1}^{p}\gamma\mu'(X_{k-1}^x)\right),$
we get
\[
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]\leq\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \left(\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4)\right)^{2},
\]
where we denote
\[
\zeta_{l,j}(u)=\mathsf{E}\left[\rme^{-u\gamma\sum_{k=l}^{j}\mu'(X^x_{k-1})}\right],\quad u>0.
\]
Note that due to our assumptios, $\mu(x) = \mu_0(x) + \Delta(x)$ with $\mu^{\prime}_0(x) < -m$. Note that $\left|\Delta^{\prime}(x)\right| \leq M^{\prime}$, hence
\begin{equation}
\label{eq:zeta}
\zeta_{l,j}(u) \leq \mathsf{E}\left[\rme^{-u\gamma\sum_{k=l}^{j}\mu_0'(X^x_{k-1})}\right]\rme^{u\gamma(j-l+1)M^{\prime}}
\end{equation}
To bound the first exponential moment in \ref{eq:zeta}, we apply
Lemma~\ref{lem:exp_bound} in Appendix~\ref{sec:appendix}. It implies that
\begin{align*}
\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{(j-l+1)\gamma\left(4u^2\kappa^{-2} \|\mu_0^{\prime}\|^2_{\text{Lip}} + M^2 + M^{\prime} + C_1u\sqrt{\gamma} -  u\pi_{\mu_0}(\mu_0')\right)\right\}, \quad \gamma(j-l+1) \geq C_0 \\
\exp\left\{uC_0(\max\limits_{y}|\mu_0'(y)| + M^{\prime})\right\}, \quad 0 \leq \gamma(j-l+1) \leq C_0	
\end{cases}
\end{align*}
with $C_0 = \left(\frac{4}{m+L} + \kappa^{-1}\right)$, measure $\pi_{\mu_{0}}$ and constants $C(x)$ and $C_1$ given in Lemma~\ref{lem:exp_bound} . Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R})$ and $\pi^{\prime}(x) = -\frac{1}{2}\pi(x)\mu(x)$, we have
\[
\pi(\mu')=\int\mu'(x)\pi(x)\,dx=-\int\mu(x)\pi'(x)\,dx=\frac{1}{2}\int\mu^{2}(x)\pi(x)\,dx>0
\]
Hence the following bound follows
\begin{align*}
\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{-\gamma(j-l)\alpha\right\}, \quad \gamma(j-l) \geq C_0 \\
\exp\left\{uC_0\max\limits_{y}|\mu'(y)|\right\}, \quad 0 \leq \gamma(j-l) \leq C_0	
\end{cases}
\end{align*}
with the quantity $ \alpha = \min(...,...) > 0$. Hence we obtain
\begin{align*}
\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4) &\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + C(x)\sum_{p=j+1}^{\infty} \rme^{-\gamma(p-j)\alpha}
\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{- \gamma\alpha}}
\end{align*}
and
\begin{align*}
\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{- \gamma\alpha}}
\end{align*}
Thus the final bound follows:
\begin{align*}
\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X^x_{p}\right)\right] \leq \gamma^{-1} C_K C(x)
\end{align*}
 with $C_K$  depending neither on $q$ nor on $\gamma$. The proof is completed.
\end{proof}

@@

\begin{lem}
\label{lem:covariance}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Let $\tilde{f}(x) = f(x) - \pi(f)$. Then
\begin{equation}
\label{eq:cov_v_ergodic}
\PCov_x \left[f(X_k),f(X_{k+s})\right] \leq C^2V^2(x)\rho^{k+s} + C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x) + C\pi(V)\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}
\end{equation}
\end{lem}
\begin{proof}
Note that
\[
\PCov_x \left[f(X_k),f(X_{k+s})\right] = \PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] + \left[\pi(f) - \PE_xf(X_s)\right]\left[\PE_xf(X_k) - \pi(f)\right]
\]
Due to $V-$ergodicity,
\[
\left|\left[\pi(f) - \PE_xf(X_s)\right]\left[\PE_xf(X_k) - \pi(f)\right]\right| \leq C^2V^2(x)\rho^{k+s}
\]
To bound the first term note that
\begin{multline*}
\left|\PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] - \PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right|
\\
\leq \int\limits_{\rset^d}\left|\PE_y\left[\tilde{f}(X_0)\tilde{f}(X_{s})\right]\right| |\P^k(x,dy)-\pi(dy)|
\leq C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x),
\end{multline*}
where the last inequality is due to lemma~\ref{lem:fp_covariance}, which implies \eqref{eq:cov_v_ergodic}.
\end{proof}
Now we state and prove the main result of this section on the variance bound for the estimator $\pi_N^n(f)$ for $V-$geometrically ergodic Markov chain.
\begin{lem}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Assume also that $n(1-\rho^{1/2}) > 1$. Then
\begin{equation}
\label{eq:vanilla_var}
\PVar_x \left[ \pi_{N}^n(f) \right] \leq \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})} + \mathcal{O}(\frac{1}{n^2(1-\rho^{1/2})})
\end{equation}
\end{lem}
\begin{proof}
Note that
\begin{eqnarray*}
\PVar_x \left[ \frac{1}{n}\sum\limits_{k=N+1}^{N+n}f(X_k) \right] = \underbrace{\frac{1}{n^2}\sum\limits_{k=N+1}^{N+n}\PVar_x \left[ f(X_k) \right]}_{S_1} + \underbrace{\frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\sum\limits_{s=1}^{n-k-1}\PCov_x\left[f(X_k),f(X_{k+s})\right]}_{S_2}
\end{eqnarray*}
Now we bound first and second sum using lemma~\ref{lem:covariance}:
\[
S_1 \leq \frac{1}{n}C\pi(V)\|\tilde{f}\|_{V^{1/2}} + \frac{\rho^N C^2\left(V^2(x) + V(x)\|\tilde{f}\|_{V^{1/2}}\right)}{n^2(1-\rho)}
\]
\begin{eqnarray*}
S_2 & \leq & \frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\left[C^2V^2(x)\rho^{k+1}\frac{1}{1-\rho} + C^2V(x)\rho^{k+1/2}\frac{1}{1-\rho^{1/2}} + C\pi(V)\frac{1}{1-\rho^{1/2}}\|\tilde{f}\|_{V^{1/2}}\right] \leq \\
& \leq & \frac{2\rho^NC^2V^2(x)}{n^2(1-\rho)^2} + \frac{2\rho^NC^2V(x)}{n^2(1-\rho)(1-\rho^{1/2})} + \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})}
\end{eqnarray*}
Hence \eqref{eq:vanilla_var} follows.
\end{proof}

@@

 \pi(y) = \frac{\exp\left\{\int\limits_{0}^{y}\mu_{0}(x)\,dx\right\}}{\int\limits_{-\infty}^{+\infty}\exp\left\{\int\limits_{0}^{y}\mu_{0}(x)\,dx\right\}\,dy}

>>>>

Now, from the H\"older inequality, we obtain (with $\|X\|_p = (\mathsf{E}X^p)^{\frac{1}{p}}$)
\begin{align*}
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=l}^{j}\alpha_{k}\left( \sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right] & \leq\sum_{j=1}^{q}\left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left\Vert \sum_{p=j+1}^{q+1} \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}^{2}\\
 & \leq\sum_{j=1}^{q} \left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left(\sum_{p=j+1}^{q+1}\left\Vert \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}\right)^{2}.
\end{align*}
Now using the fact that $\prod\limits_{k=j+1}^{p}\alpha_{k} \leq \exp\left(-\sum\limits_{k=j+1}^{p}\gamma\mu'(X_{k-1}^x)\right),$
we get
\[
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]\leq\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \left(\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4)\right)^{2},
\]
where we denote
\[
\zeta_{l,j}(u)=\mathsf{E}\left[\rme^{-u\gamma\sum_{k=l}^{j}\mu'(X^x_{k-1})}\right],\quad u>0.
\]
Note that due to our assumptios, $\mu(x) = \mu_0(x) + \Delta(x)$ with $\mu^{\prime}_0(x) < -m$. Note that $\left|\Delta^{\prime}(x)\right| \leq M^{\prime}$, hence
\begin{equation}
\label{eq:zeta}
\zeta_{l,j}(u) \leq \mathsf{E}\left[\rme^{-u\gamma\sum_{k=l}^{j}\mu_0'(X^x_{k-1})}\right]\rme^{u\gamma(j-l+1)M^{\prime}}
\end{equation}
To bound the first exponential moment in \ref{eq:zeta}, we apply
Lemma~\ref{lem:exp_bound} in Appendix~\ref{sec:appendix} to $f(x) = -\mu^{\prime}_0(x)$. It implies that
\begin{align*}
\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{(j-l+1)\gamma\left(4u^2\kappa^{-2} \|\mu_0^{\prime}\|^2_{\text{Lip}} + M^2 + M^{\prime} + C_1u\sqrt{\gamma} +  u\pi_{\mu_0}(\mu_0')\right)\right\}, \quad \gamma(j-l+1) \geq C_0 \\
\exp\left\{uC_0(\max\limits_{y}|\mu_0'(y)| + M^{\prime})\right\}, \quad 0 \leq \gamma(j-l+1) \leq C_0	
\end{cases}
\end{align*}
with $C_0 = \left(\frac{4}{m+L} + \kappa^{-1}\right)$, measure $\pi_{\mu_{0}}$ and constants $C(x)$ and $C_1$ given in Lemma~\ref{lem:exp_bound} . Furthermore, since $\mu_0\pi_{\mu_0}',\mu_0'\pi_{\mu_0}\in L^{1}(\rset)$ and $\pi_{\mu_0}^{\prime}(x) = \pi_{\mu_0}(x)\mu_0(x)$, we have
\[
\pi_{\mu_0}(\mu_0')=\int\mu_0'(x)\pi_{\mu_0}(x)\,dx=-\int\mu_{\pi_0}(x)\pi_{\mu_0}'(x)\,dx= -\int\mu_0^{2}(x)\pi_{\mu_0}(x)\,dx < 0
\]
Hence the following bound follows
\begin{align*}
\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{\gamma(j-l+1)\alpha\right\}, \quad \gamma(j-l+1) \geq C_0 \\
\exp\left\{(\max\limits_{y}|\mu_0'(y)| + M^{\prime})\right\}, \quad 0 \leq \gamma(j-l+1) \leq C_0
\end{cases}
\end{align*}
with the quantity 
\[
\alpha = \max(64\kappa^{-2} \|\mu_0^{\prime}\|^2_{\text{Lip}} + M^2 + M^{\prime} + 4C_1\sqrt{\gamma} +  4\pi_{\mu_0}(\mu_0'), 16\kappa^{-2} \|\mu_0^{\prime}\|^2_{\text{Lip}} + M^2 + M^{\prime} + 2C_1\sqrt{\gamma} +  2\pi_{\mu_0}(\mu_0')) < 0
\] 
Hence we obtain
\begin{align*}
\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4) &\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + C(x)\sum_{p=j+1}^{\infty} \rme^{\gamma(p-j)\alpha}
\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{\gamma\alpha}}
\end{align*}
and
\begin{align*}
\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{\gamma\alpha}}
\end{align*}
Thus the final bound follows:
\begin{align*}
\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X^x_{p}\right)\right] \leq \gamma^{-1} C_K C(x)
\end{align*}
 with $C_K$  depending neither on $q$ nor on $\gamma$. The proof is completed.
\end{proof}

@@

\begin{lem}
\label{lem:covariance}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Let $\tilde{f}(x) = f(x) - \pi(f)$. Then
\begin{equation}
\label{eq:cov_v_ergodic}
\PCov_x \left[f(X_k),f(X_{k+s})\right] \leq C^2V^2(x)\rho^{k+s} + C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x) + C\pi(V)\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}
\end{equation}
\end{lem}
\begin{proof}
Note that
\[
\PCov_x \left[f(X_k),f(X_{k+s})\right] = \PE_x\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right] + \left[\pi(f) - \PE_xf(X_{k+s})\right]\left[\PE_xf(X_k) - \pi(f)\right]
\]
Due to $V-$ergodicity,
\[
\left|\left[\pi(f) - \PE_xf(X_{k+s})\right]\left[\PE_xf(X_k) - \pi(f)\right]\right| \leq C^2V^2(x)\rho^{2k+s}
\]
To bound the first term note that
\begin{multline*}
\left|\PE_x\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right] - \PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right| = \left|\PE_x\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right] - \PE_\pi\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right]\right| \leq
\\
\leq \int\limits_{\rset^d}\left|\PE_y\left[\tilde{f}(X_0)\tilde{f}(X_{s})\right]\right| |\P^k(x,dy)-\pi(dy)|
\leq C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x),
\end{multline*}
where the last inequality is due to lemma~\ref{lem:fp_covariance}, which implies \eqref{eq:cov_v_ergodic}.
\end{proof}
Now we state and prove the main result of this section on the variance bound for the estimator $\pi_N^n(f)$ for $V-$geometrically ergodic Markov chain.
\begin{lem}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Assume also that $n(1-\rho^{1/2}) > 1$. Then
\begin{equation}
\label{eq:vanilla_var}
\PVar_x \left[ \pi_{N}^n(f) \right] \leq \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})} + \mathcal{O}(\frac{1}{n^2(1-\rho^{1/2})})
\end{equation}
\end{lem}
\begin{proof}
Note that
\begin{eqnarray*}
\PVar_x \left[ \frac{1}{n}\sum\limits_{k=N+1}^{N+n}f(X_k) \right] = \underbrace{\frac{1}{n^2}\sum\limits_{k=N+1}^{N+n}\PVar_x \left[ f(X_k) \right]}_{S_1} + \underbrace{\frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\sum\limits_{s=1}^{n-k-1}\PCov_x\left[f(X_k),f(X_{k+s})\right]}_{S_2}
\end{eqnarray*}
Now we bound first and second sum using lemma~\ref{lem:covariance}:
\[
S_1 \leq \frac{1}{n}C\pi(V)\|\tilde{f}\|_{V^{1/2}} + \frac{\rho^N C^2\left(V^2(x) + V(x)\|\tilde{f}\|_{V^{1/2}}\right)}{n^2(1-\rho)}
\]
\begin{eqnarray*}
S_2 & \leq & \frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\left[C^2V^2(x)\rho^{k+1}\frac{1}{1-\rho} + C^2V(x)\rho^{k+1/2}\frac{1}{1-\rho^{1/2}} + C\pi(V)\frac{1}{1-\rho^{1/2}}\|\tilde{f}\|_{V^{1/2}}\right] \leq \\
& \leq & \frac{2\rho^NC^2V^2(x)}{n^2(1-\rho)^2} + \frac{2\rho^NC^2V(x)}{n^2(1-\rho)(1-\rho^{1/2})} + \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})}
\end{eqnarray*}
Hence \eqref{eq:vanilla_var} follows.
\end{proof}

@@

 \pi_{\mu_0}(y) = \frac{\exp\left\{\int\limits_{0}^{y}\mu_{0}(x)\,dx\right\}}{\int\limits_{-\infty}^{+\infty}\exp\left\{\int\limits_{0}^{y}\mu_{0}(x)\,dx\right\}\,dy}
